Slovene NLTK Tagger

### 2012-01-08 paper/paper.tex
Fixed some errors.

\section{Introduction} %NIKICC
First, let us say a few words about the structure of the paper. In this chapter, we provide a breif theorethical description of imlemented taggers. Next we will describe the related work in section \ref{relatedWork}, which was the base of our project. In section \ref{implementation} the proces of implementaion itself is explained. The example, how to use the tagger is provided, along with some discussion on encoding problems. Finally in section \ref{results} is the evaluation of accuracy and speed.
\par
Now we begin with the description of implemented taggers.

%NIKICC
\subsection[Trigram tagger]{Trigram tagger\footnote{Definition and images summerized from \cite{NLTKBOOK}}}
To understant how \textit{trigram} works, we should first understand \textit{unigram}.
Unigram taggers are based on a simple statistical algorithm: \textit{for each token, assign the tag that is most likely for that particular token}.
In the case of tagging with unigram tagger, we only consider the current token, in isolation from any larger context. The best we can do is tag each word with its \textit{a priori} most likely tag.
\par
An n-gram tagger is a generalization of a unigram\footnote{Unigram could also be reffered to as 1-gram} tagger whose context is the current word together with the part-of-speech tags of the $n-1$ preceding tokens, as shown in figure \ref{fig:trigram}.

@@

The tag to be chosen, $t_{n}$, is circled, and the context is shaded in grey.
In the example of an n-gram tagger shown in figure \ref{fig:trigram}, we have $n=3$; that is, we consider the tags of the two preceding words in addition to the current word.
An n-gram tagger picks the tag that is most likely in the given context.
\par
Altrough we use the name trigram all throughtout this paper, we acctually refer to sequential tagger, which consists of
\textit{affix}\footnote{The Affix tagger learns prefix and suffix patterns to determine the part of speech tag for word.}, \textit{unigram}, \textit{bigram} and \textit{trigram} taggers in this order respectively. First we build affix, which is used as ad \textit{backoff-tagger} for unigram. Unigram is afterwards used as a backoff-tagger for bigram, which is finally used for trigram.

@@

Despite obvious differences between Brill tagger and n-gram tagger there is one similarity. Both of them use \textit{supervised learning} method, since we require annotated training data to make sure whether the tagger's guess is a mistake or not.

@@

\subsection[Naive Bayes classifier]{Naive Bayes classifier\footnote{Definition and images summerized from \cite{NLTKBOOK}}}
Both Brill and Trigram are sequential backoff taggers. Meanwhile, Naive Bayes is \textit{classifier based tagger}.
Classification is a process where, for a given output, we have to choose the correct class. Basically, it is an  procedure for assigning  input data into one of a given number of categories.

Examples of classification tasks are:
\begin{itemize}
\item whether an email is spam or not.
\item diagnosis to a given patient as described by observed characteristics of the patient
\end{itemize}
The basic classification task has a number of interesting variants. For example, in multi-class classification, each instance may be assigned multiple labels; in open-class classification, the set of labels is not defined in advance; and in sequence classification, a list of inputs are jointly classified.

@@

Features make their contribution to the  decision by voting against those labels, that don't occur with this feature. Likelihood score is therefore reduced for each label by multiplying it with the probability that an input value with that label would have that feature. Suppose that the word occurs ran occurs in  12\% of the sports documents, 10\% of the murder mystery documents, and 2\% of the automotive documents. The likelihood score for the sports label will be multiplied by 0.12, the likelihood score for the murder mystery label will be multiplied by 0.1, and the likelihood score for the automotive label will be multiplied by 0.02. The overall effect will be to slightly reduce the score of the sports label, reduce the murder mystery label (more then sports label) and significantly reduce the automotive label. Figure \ref{fig:naive-bayes-bargraph} illustrates this example.

@@

\section{Related work}
\label{relatedWork}
\subsection{Nltk-trainer} %NIKICC
The code for building the tagger was taken from \textit{Nltk-trainer}  project \cite{nltk-trainer}, whose author is Jacob Perkins.
Its an open-source project hosting on Github, with moto to \textit{Train NLTK objects with zero code}.
We almost exclusively used the script \texttt{train\_tagger.py}, which has the ability to generate NLTK taggers. This project made it a lot easier for us to train the taggers on Slovenian corpuses.

\subsection{JOS corpus} %BANIČ
The basic component for building Slovene tagger is \textit{JOS corpus} from JOS project \cite{JOS}. It stands for: \textit{Jezikoslovno Označevanje Slovenskega jezika} (Linguistic annotation of Slovenian language).
It contains collections of various text in Slovenian language with annotation.
For our project we used jos1M corpus that contain 1 million words with it's appropriate lemmas and morphosyntactic descriptions.
The JOS is compatible with Slovene\textit{ MULTEXT-East morphosyntactic specifications Version 4}\cite{MULTEXT-East}.
It basic purpose is to define word classes.
For each class there are various attributes and their appropriate values, which they can be mapped into morphosyntactic descriptions (i.e. MSD).
The structure of JOS corpora is in  \textit{Extensible Markup Language (XML)}.

@@

We can easily manipulate XML files in various programming languages (e.g. Python with xml.dom.minidom).
This is mandatory because \textit{nltk-trainer}\cite{nltk-trainer} expects special form of input file (.pos file).

@@

The trainer will train the tagger properly, if in .pos file:
\begin{itemize}
\item The words and their MSD are separated with slash.
\item Every punctuation mark is separated with slash and then follows the same punctuation mark.
\item Words and punctuation marks are between space.
\end{itemize}

@@

\subsection{Training the tagger} %NIKICC
As mentioned above, the script \texttt{train\_tagger.py} was used for this purpose. Its mandatory argument is a corpus in .pos format, which generation is described in \ref{Corpus transformation}. Beside this, there are many optional arguments. Here we will describe basics, for further description run script help with
\begin{lstlisting}
train_tagger.py --help
\end{lstlisting}
\par
One is wheather to generate a Sequential Tagger, a Brill Tagger or a Classifier Based Tagger. For Sequential we can then select sequential backoff algorithm. This can be any combination of the following letters:
\begin{itemize}
\item a: AffixTagger
\item u: UnigramTagger
\item b: BigramTagger
\item t: TrigramTagger
\end{itemize}
The default is \textit{aubt}, which was used for our Trigram tagger, but you can set this to the empty string to not train a sequential backoff tagger. The Brill Tagger is trained in front of the other tagger, which in our implementation was the Trigram tagger. When building classifier based algorithm, we must select the classifier or the sequence of classifier. Following are supported: NaiveBayes, DecisionTree, Maxent, GIS, IIS, CG, BFGS, Powell, LBFGSB, Nelder-Mead, MEGAM and TADM. We should mention, that the time spent for training classifier based tagger, is much larger than, the time for a sequential or brill.
\par
Script also has the possibility to set default tag. We used \textit{-Neznan-}, which is Slovene word for \textit{-unknown-}.
\par
Last argument, that should be mentioned, is wheather to evaluate the tagger or not and the procentage of sentaces to be used. The default value is 1, which means that the \textit{same} sentances are first used for training the tagger and then for evaluation. We recommend this argument to be set below 1. If so, the right approach is used.
Sentances are devided into two groups, one in used for training and second for evaluation. \textit{Fraction} is the procentage of sentances assigned to training group. 
This was used for evaluating constructed taggers. For results of this evaluation see section \ref{results}).

@@

As you can see, there is the problem with the last dot. The solution we used is described below.

@@

We also implemented a function that returns two description dictionaries for MSDs with attribute=value expansions in Slovene and English. This way user can view which part-of-speech words belong to in Slovene or English. Following example demonstrates the expected output:
\\\\
\textit{( Lep  |  PPNMEIN ) - pridevnik \\
	( je  |  GP-STE-N ) - glagol\\
	( dan  |  SOMEI ) - samostalnik\\
	( ,  |  , ) - ni razlage\\
	( vse  |  ZC-SEI ) - zaimek\\
	( diši  |  GGNSTE ) - glagol\\
	( že  |  L ) - členek\\
	( po  |  DM ) - predlog\\
	( pomladi  |  SOZEM ) - samostalnik\\
	( !  |  ! ) - ni razlage}\\

@@

The first output was creted with function \textit{prettyPrintWithDescription}, full output can be created with \textit{prettyPrintWithFullDescription}. Both function are defined in \texttt{example.py}
\par
But there is an issue. Because Slovenian language contain letters with caron, the output will have unicode symbols for them. In the previous examples, the output is already corrected. Solution is described in the following section.

@@

\section{Results} %NIKICC
\label{results}
For all constructed taggers, we here provide their accuracy. The results were obtained by repeating the proces of building the tagger with various evaluation arguments (see script \texttt{evaluateTaggers.sh}). See table \ref{tab:evaluation} and figure \ref{fig:evaluation}.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.5\textwidth]{../evaluation/graph.pdf} 
\end{center}
\caption{Accuracy evaluation results.}
\label{fig:evaluation}
\end{figure}
\par
To conclude, there are no major differences in tagging accuracy. The Brill and the Naive Bayes are about 2\% better than Trigram.
\par
However, during this evaluation we discover an other parameter, which should be taken into account.
Under the assumption that most of the time was actually spent for tagging the sentences (besides just testing for equality and percentage calculation are needed), we noticed that the time for evaluations differ drastically.
The results of resarching time consumption are listed in table \ref{tab:evaluation_speed} and graphically represented on figure \ref{fig:evaluation_speed} (see script \texttt{evaluateTaggersSpeed.py}).

@@

The difference here is enormous. Brill and Trigram run at almost the same speed, while Naive Bayes runs extremely slower. This is also one of the reason why we prefer usig the Brill tagger.

\section{Conclusion} %BANIČ
To conclude our paper, tagging in Slovene language is practical with proper use of Natural Language Toolkit and \textit{nltk-trainer}. Time spent in seconds for tagging different numbers of words, show that Brill and Trigram ran at the same speed, while Naive Bayes ran drasticly slower. There are no major differences in tagging accuracy. However Brill and the Naive Bayes are about 2\% better than Trigram.
\par
Besides these three taggers we built, others can be built without any bigger modification. All that is needed is to set some arguments when calling the script \texttt{train\_tagger.py}. The main reason we built only these three is computing power. These three are the only one we could built and evaluate in some resonable time.
\par
Our whole project is publicly available on github \cite{CODE}. We also contacted the NLTK developers group and will arrange our taggers to become the part of NLTK.

>>>>

\section{Introduction} %NIKICC
First, let us say a few words about the structure of the paper. In this chapter, we provide a brief theoretical description of implemented taggers. Next we will describe the related work in section \ref{relatedWork}, which was the base of our project. In section \ref{implementation} the process of implementation itself is explained. The example, how to use the tagger is provided, along with some discussion on encoding problems. Finally in section \ref{results} is the evaluation of accuracy and speed.
\par
Now we begin with the description of implemented taggers.

%NIKICC
\subsection[Trigram tagger]{Trigram tagger\footnote{Definition and images summarized from \cite{NLTKBOOK}}}
To understand how \textit{trigram} works, we should first understand \textit{unigram}.
Unigram taggers are based on a simple statistical algorithm: \textit{for each token, assign the tag that is most likely for that particular token}.
In the case of tagging with Unigram tagger, we only consider the current token, in isolation from any larger context. The best we can do is tag each word with its \textit{a priori} most likely tag.
\par
An n-gram tagger is a generalization of a unigram\footnote{Unigram could also be refereed to as 1-gram} tagger whose context is the current word together with the part-of-speech tags of the $n-1$ preceding tokens, as shown in figure \ref{fig:trigram}.

@@

The tag to be chosen, $t_{n}$, is circled, and the context is shaded in grey.
In the example of an n-gram tagger shown in figure \ref{fig:trigram}, we have $n=3$; that is, we consider the tags of the two preceding words in addition to the current word.
An n-gram tagger picks the tag that is most likely in the given context.
\par
Although we use the name trigram all throughout this paper, we actually refer to sequential tagger, which consists of
\textit{affix}\footnote{The Affix tagger learns prefix and suffix patterns to determine the part of speech tag for word.}, \textit{unigram}, \textit{bigram} and \textit{trigram} taggers in this order respectively. First we build affix, which is used as ad \textit{backoff-tagger} for unigram. Unigram is afterwards used as a backoff-tagger for bigram, which is finally used for trigram.

@@

Despite obvious differences between Brill tagger and n-gram tagger there is one similarity. Both of them use \textit{supervised learning} method, since we require annotated training data to make sure whether the taggers guess is a mistake or not.

@@

\subsection[Naive Bayes classifier]{Naive Bayes classifier\footnote{Definition and images summarized from \cite{NLTKBOOK}}}
Both Brill and Trigram are sequential backoff taggers. Meanwhile, Naive Bayes is \textit{classifier based tagger}.
Classification is a process where, for a given output, we have to choose the correct class. Basically, it is an  procedure for assigning input data into one of a given number of categories.

Examples of classification tasks are:
\begin{itemize}
\item whether an email is spam or not.
\item diagnosis to a given patient as described by observed characteristics of the patient.
\end{itemize}
The basic classification task has a number of interesting variants. For example, in multi-class classification, each instance may be assigned multiple labels; in open-class classification, the set of labels is not defined in advance; and in sequence classification, a list of inputs are jointly classified.

@@

Features make their contribution to the decision by voting against those labels, that don't occur with this feature. Likelihood score is therefore reduced for each label by multiplying it with the probability, that an input value with that label would have that feature. Suppose that the word \textit{ran} occurs in  12\% of the sports documents, 10\% of the murder mystery documents, and 2\% of the automotive documents. The likelihood score for the sports label will be multiplied by 0.12, the likelihood score for the murder mystery label will be multiplied by 0.1, and the likelihood score for the automotive label will be multiplied by 0.02. The overall effect will be, to slightly reduce the score of the sports label, reduce the murder mystery label (more then sports label) and significantly reduce the automotive label. Figure \ref{fig:naive-bayes-bargraph} illustrates this example.

@@

\section{Related work}
\label{relatedWork}
\subsection{Nltk-trainer} %NIKICC
The code for building the tagger was taken from \textit{Nltk-trainer}  project \cite{nltk-trainer}, whose author is Jacob Perkins.
Its an open-source project hosting on Github, with moto to \textit{Train NLTK objects with zero code}.
We almost exclusively used the script \texttt{train\_tagger.py}, which has the ability to generate NLTK taggers. This project made it a lot easier for us to train the taggers on Slovenian corpora.

\subsection{JOS corpus} %BANIČ
The basic component for building Slovene tagger is \textit{JOS corpus} from JOS project \cite{JOS}. It stands for: \textit{Jezikoslovno Označevanje Slovenskega jezika} (Linguistic annotation of Slovenian language).
It contains collections of various text in Slovenian language with annotation.
For our project we used jos1M corpus, that contain 1 million words with it's appropriate lemmas and morphosyntactic descriptions.
The JOS is compatible with Slovene\textit{ MULTEXT-East morphosyntactic specifications Version 4}\cite{MULTEXT-East}.
It basic purpose is to define word classes (i.e. part-of-speech).
For each class there are various attributes and their appropriate values, which they can be mapped into morphosyntactic descriptions (i.e. MSD).
The structure of JOS corpora is in  \textit{Extensible Markup Language (XML)}.

@@

We can easily manipulate XML files in various programming languages (e.g. Python with xml.dom.minidom).
This is mandatory, because \textit{nltk-trainer}\cite{nltk-trainer} expects special form of input file (.pos file).

@@

The trainer will train the tagger properly, if .pos file consists off:
\begin{itemize}
\item The words and their MSD are separated with slash.
\item Every punctuation mark is separated with slash and then follows the same punctuation mark.
\item Words and punctuation marks are between space.
\end{itemize}

@@

\subsection{Training the tagger} %NIKICC
As mentioned above, the script \texttt{train\_tagger.py} was used for this purpose. Its mandatory argument is a corpus in .pos format, which generation is described in \ref{Corpus transformation}. Beside this, there are many optional arguments. Here we will describe basics, for further description run script help with:
\begin{lstlisting}
train_tagger.py --help
\end{lstlisting}
\par
One is whether to generate a Sequential Tagger, a Brill Tagger or a Classifier Based Tagger. For Sequential we can then select sequential backoff algorithm. This can be any combination of the following letters:
\begin{itemize}
\item a: AffixTagger
\item u: UnigramTagger
\item b: BigramTagger
\item t: TrigramTagger
\end{itemize}
The default is \textit{aubt}, which was used for our Trigram tagger, but you can set this to the empty string to not train a sequential backoff tagger. The Brill tagger is trained in front of the other tagger, which in our implementation was the Trigram tagger. When building classifier based algorithm, we must select the classifier or the sequence of classifier. Following are supported: NaiveBayes, DecisionTree, Maxent, GIS, IIS, CG, BFGS, Powell, LBFGSB, Nelder-Mead, MEGAM and TADM. We should mention, that the time spent for training classifier based tagger, is much larger than, the time for a sequential or brill.
\par
Script also has the possibility to set default tag. We used \textit{-Neznan-}, which is Slovene word for \textit{-unknown-}.
\par
Last argument, that should be mentioned, is whether to evaluate the tagger or not and the percentage of sentences to be used. The default value is 1, which means that the \textit{same} sentences are first used for training the tagger and then for evaluation. We recommend this argument to be set below 1. If so, the right approach is used.
Sentences are divided into two groups, one in used for training and second for evaluation. \textit{Fraction} is the percentage of sentences assigned to training group. 
This was used for evaluating constructed taggers. For results of this evaluation see section \ref{results}).

@@

As you can see, there is a problem with the last token. The solution we used is described below.

@@

We also implemented a function that returns two description dictionaries for MSD's with attribute=value expansions in Slovene and English. This way user can view which part-of-speech words belong to in Slovene or English. Following example demonstrates the expected output:
\\\\
\textit{( Lep  |  PPNMEIN ) - pridevnik \\
	( je  |  GP-STE-N ) - glagol\\
	( dan  |  SOMEI ) - samostalnik\\
	( ,  |  , ) - ni razlage\\
	( vse  |  ZC-SEI ) - zaimek\\
	( diši  |  GGNSTE ) - glagol\\
	( že  |  L ) - členek\\
	( po  |  DM ) - predlog\\
	( pomladi  |  SOZEM ) - samostalnik\\
	( !  |  ! ) - ni razlage}\\

@@

The first output was created with function \textit{prettyPrintWithDescription}, full output can be created with \textit{prettyPrintWithFullDescription}. Both function are defined in \texttt{example.py}
\par
But there is an issue. Because Slovenian language contain letters with caron, the output will have unicode symbols for them. In the previous examples, the output is already corrected. Solution is described in the following section.

@@

\section{Results} %NIKICC
\label{results}
For all constructed taggers, we here provide their accuracy. The results were obtained by repeating the process of building the tagger with various evaluation arguments (see script \texttt{evaluateTaggers.sh}). See table \ref{tab:evaluation} and figure \ref{fig:evaluation}.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.5\textwidth]{../evaluation/graph.pdf} 
\end{center}
\caption{Accuracy evaluation results.}
\label{fig:evaluation}
\end{figure}
\par
To conclude, there are no major differences in tagging accuracy. The Brill and the Naive Bayes are about 2\% better than Trigram.
\par
However, during this evaluation we discover an other parameter, which should be taken into account.
Under the assumption that most of the time was actually spent for tagging the sentences (besides just testing for equality and percentage calculation are needed), we noticed that the time for evaluations differ drastically.
The results of researching time consumption are listed in table \ref{tab:evaluation_speed} and graphically represented on figure \ref{fig:evaluation_speed} (see script \texttt{evaluateTaggersSpeed.py}).

@@

The difference here is enormous. Brill and Trigram run at almost the same speed, while Naive Bayes runs extremely slower. This is also one of the reason why we prefer using the Brill tagger.

\section{Conclusion} %BANIČ
To conclude our paper, tagging in Slovene language is practical with proper use of Natural Language Toolkit and \textit{nltk-trainer}. Time spent in seconds for tagging different numbers of words, show that Brill and Trigram ran at the same speed, while Naive Bayes ran drastically slower. There are no major differences in tagging accuracy. However Brill and the Naive Bayes are about 2\% better than Trigram.
\par
Besides these three taggers we built, others can be built without any bigger modification. All that is needed is to set some arguments when calling the script \texttt{train\_tagger.py}. The main reason we built only these three is computing power. These three are the only one we could built and evaluate in some reasonable time.
\par
Our whole project is publicly available on github \cite{CODE}. We also contacted the NLTK developers group and will arrange our taggers to become the part of NLTK.
